---
title: "Exercise 1"
author: "Ivan Quirante González"
date: "2026-09-17"
output: html_document
---

# SNP dataset

### 0. Loading the data

```{r}
setwd("G:/My Drive/University/Masters/MDS/q3/BSG/sge-exercises")

raw <- read.table("TSICHR22RAW.raw", header = TRUE)
X   <- raw[, 7:ncol(raw)]                 # only the genetic information
```

Each cell corresponds to the number of B alleles, 0=AA, 1=AB, 2 = BB.

### 1. How many variants are there in this database? What percentage of the data is missing?

```{r}
ncol(X)
100*mean(is.na(X))
```

This database has 20,649 variants an a \~0.20% of missing data.

### 2. Calculate the percentage of monomorphic variants. Exclude all monomorphics from the database for all posterior computations of the practical. How many variants do remain in your database?

A **monomorphic** variant is one where everybody in the sample carries the same allele. There's no variation at that position, so it's not really a polymorphism in this dataset.

```{r}
nB   <- colSums(X, na.rm = TRUE)          # copies of allele B
nTot <- 2 * colSums(!is.na(X))            # total alleles typed
mono <- nB == 0 | nB == nTot              # only one allele present
sum(mono)
100 * mean(mono)                          # % monomorphic
X <- X[, !mono]                           # remove them
ncol(X)                                   # variants remaining
```

We can see that an 11.46% of the variants are monomorphic. If we remove them we end up with 18283 variants.

### 3. Report the genotype counts and the minor allele count of polymorphism `rs8138488_C`, and calculate the MAF of this variant.

```{r}
g  <- X[, "rs8138488_C"]
gc <- table(factor(g, levels = 0:2, labels = c("AA", "AB", "BB")))
gc                                        # genotype counts
count_A <- 2 * gc[["AA"]] + gc[["AB"]]
count_B <- 2 * gc[["BB"]] + gc[["AB"]]        # B = the C allele
min(count_A, count_B)                   # minor allele count
min(count_A, count_B) / (count_A + count_B)   # MAF

```

For polymorphism `rs8138488_c`, we have 41 individuals with genotype AA, 47 with genotype AB and 14 with BB.

The minor allele is C, which is the counted one in the dataset. It has a MAF of 0.368.

### 4. Compute the minor allele frequencies (MAF) for all markers, and make a histogram of it. Does the MAF follow a uniform distribution? What percentage of the markers have a MAF below 0.05? And below 0.01? Can you explain the observed pattern? 

*Uniform distribution at [0, 0.5]. The range from 0 to 0.5 can be easily explained due to the fact that a frequency higher than 0.5 would indicate this is not the minor allele.*

```{r}
p   <- colSums(X, na.rm = TRUE) / (2 * colSums(!is.na(X)))
maf <- pmin(p, 1 - p)
hist(maf, breaks = 50, col = "steelblue",
     main = "Minor allele frequency of SNPs (chr22, TSI)",
     xlab = "Minor allele frequency", ylab = "Number of SNPs")
100 * mean(maf < 0.05)                    # % MAF < 0.05
100 * mean(maf < 0.01) 
```

No, we see instead a shape more similar to $\frac{1}{x}$ rather than a uniform distribution. This shape in population genetics is called the **site / allele frequency spectrum**, and it is characterized by a big spike at the lowest frequencies followed by a decay as we move right.

A 14.23% of markers have a MAF below 0.05, and a 4.73% are below 0.01. In this sample of 102 individuals (204 alleles), this means that 14.23% of variants have at most $0.05 \times 204 \approx 10$ copies of the minor allele, and 4.73% of the variants have at most $0.01 \times 204 \approx 2$ (singletons and doubletons) copies of the minor allele.

### 5. Calculate the observed heterozygosity $H_0$ , and make a histogram of it. What is, theoretically, the range of variation of this statistic?

```{r}
Ho <- colMeans(X == 1, na.rm = TRUE)
hist(Ho, breaks = 50, col = "darkorange",
     main = "Observed heterozygosity of SNPs",
     xlab = "Observed heterozygosity (Ho)", ylab = "Number of SNPs")
range(Ho)
```

In this dataset, $H_0 \in [0, 0.68]$ , but in theory $H_0 \in [0,1]$.

### 6. Compute for each marker its expected heterozygosity ($H_e$), where the expected heterozygosity for a bi-allelic marker is defined as $1 − \sum_{i=1}^kp_i^2$ , where $p_i^2$ is the frequency of the $i$-th allele. Make a histogram of the expected heterozygosity. What is, theoretically, the range of variation of this statistic? What is the average of $H_e$ for this database?

```{r}
He <- 1 - (p^2 + (1 - p)^2)
hist(He, breaks = 50, col = "seagreen",
     main = "Expected heterozygosity of SNPs",
     xlab = "Expected heterozygosity (He)", ylab = "Number of SNPs")
range(He)
mean(He)
```

In this dataset, $H_e \in [0.01, 0.5]$, but in theory $H_e \in [0, 0.5]$.

# STR dataset

### 1. How many individuals and how many STRs contains the database?

### 2. Write a function that determines the number of alleles for a STR. Determine the number of alleles for each STR in the database. Compute basic descriptive statistics of the number of alleles (mean, standard deviation, median, minimum, maximum).

### 3. Make a table with the number of STRs for a given number of alleles and present a barplot of the number STRs in each category. What is the most common number of alleles for an STR?

### 4. Compute the expected heterozygosity for each STR. Make a histogram of the expected heterozygosity over all STRS. Compute the average expected heterozygosity over all STRs.

### 5. Calculate also the observed heterozygosity for each STR. Plot observed against expected heterozygosity, using all STRs. What do you observe?

### 6. Compare, overall, the results you obtained for the SNP database with those you obtained for the STR database. What differences do you observe between these two types of genetic markers?
