---
title: "Practical 2: Hardy Weinberg Equilibrium"
subtitle: "Bioinformatics and Statistical Genetics (BSG)"
author:
  - "Ivan Quirante"
  - "Humbert Teruel"
  - "Anna Maria Zakreva"
date: "`r Sys.Date()`"
output: pdf_document
---

install.packages("HardyWeinberg")

```{r echo=FALSE}
library(data.table)
library(HardyWeinberg)
library(genetics)
```

## Loading dataset

```{r}
df <- fread("TSIChr22v4.raw")
geno_dt <- as.data.table(df[, 7:ncol(df)])
```

## 1. How many variants are there in this database? What percentage of the data is missing?
```{r}
n_variants <- ncol(geno_dt)

cat("Number of variants:", n_variants, "\n")

pct_missing <- mean(is.na(geno_dt)) * 100
cat("Percentage missing:", round(pct_missing, 4), "%\n")

```

## 2. Calculate the percentage of monomorphic variants. Exclude all monomorphics from the database for all posterior computations of the practical. How many variants do remain in your database?
```{r}

# Convert to matrix (faster than other options)
m <- as.matrix(geno_dt)

# For each column, count non-NA values equal to 0, 1, 2
has0 <- colSums(m == 0, na.rm = TRUE) > 0
has1 <- colSums(m == 1, na.rm = TRUE) > 0
has2 <- colSums(m == 2, na.rm = TRUE) > 0

# Monomorphic = only one of the three values appears
n_present <- has0 + has1 + has2
monomorphic <- n_present <= 1

cat("Monomorphic %:", mean(monomorphic) * 100, "\n")

geno_poly <- geno_dt[, !monomorphic, with = FALSE]

n_poly <- ncol(geno_poly)
cat("Remaining variants:", n_poly, "\n", "Removed variants:", ncol(geno_dt)-n_poly, "\n", "These polymorphic variants were used for all subsequent analyses.")
```

## 3. Extract polymorphism rs587756191 T from the data, and determine its genotype counts. Apply a chi-square test for Hardy-Weinberg equilibrium, with and without continuity correction. Also try an exact test, and a permutation test. You can use thee functions HWChisq, HWExact and HWPerm for this purpose. Do you think this variant is in equilibrium? Argue your answer
```{r}
"rs587756191_T" %in% colnames(geno_poly)

x <- geno_poly[["rs587756191_T"]]

# Genotype counts
g0 <- sum(x == 0, na.rm = TRUE)
g1 <- sum(x == 1, na.rm = TRUE)
g2 <- sum(x == 2, na.rm = TRUE)


counts <- c(AA = g0, AB = g1, BB = g2)
print(counts)

# Chi-Square Test
# Without continuity correction
HWChisq(counts, cc = 0)

# With continuity correction
HWChisq(counts, cc = 0.5)

# Exact Test
HWExact(counts)

# Permutation Test
HWPerm(counts)


```


The variant **rs587756191_T is in Hardy-Weinberg equilibrium**.


## 4. Determine the genotype counts for all polymorphic variants, and store them in a p × 3 matrix.
```{r}
m <- as.matrix(geno_poly)

# Count genotypes per variant
n_AA <- colSums(m == 0, na.rm = TRUE)   
n_AB <- colSums(m == 1, na.rm = TRUE)   
n_BB <- colSums(m == 2, na.rm = TRUE)   

# p × 3 matrix
genotype_counts <- cbind(AA = n_AA, AB = n_AB, BB = n_BB)

# Check dimensions
dim(genotype_counts)

# First few rows
head(genotype_counts)
```

## 5. Apply an exact test for Hardy-Weinberg equilibrium to each SNP. You can use function HWExactStats for fast computation. What is the percentage of significant SNPs (use $alpha$ = 0.05)? Is this the number of markers that you would expect to be out of equilibrium by the effect of chance alone?
```{r}
#Exact test for all SNPs
hwe_results <- HWExactStats(genotype_counts)

# Significant at alpha = 0.05
significant <- hwe_results < 0.05
n_significant <- sum(significant, na.rm = TRUE)
pct_significant <- mean(significant, na.rm = TRUE) * 100

cat("Number of significant SNPs:", n_significant, "\n")
cat("Percentage of significant SNPs:", round(pct_significant, 2), "%\n")
cat("Expected by chance (5%):", round(0.05 * length(hwe_results), 0), "\n")
```

## Conclusion


The observed percentage (2.77%) is lower than the expected 5%, meaning there are fewer deviations from HWE than would be expected by chance. This suggests the data are relatively clean.
## 6.  Which SNP is most significant according to the exact test results? Give its genotype counts. In which sense is this genotypic composition unusual?
```{r}
# Find the minimum p-value
min_p <- min(hwe_results, na.rm = TRUE)
cat("Minimum p-value:", min_p, "\n")

# Which SNP has it
most_sig <- which.min(hwe_results)

cat("Most significant SNP:", colnames(geno_poly)[most_sig], "\n")

genotype_counts[most_sig, ]
```

## Most Significant SNP

The most significant SNP according to the exact test has genotype counts of 56 AA, 0 AB, and 51 BB.

A complete absence of heterozygotes is biologically implausible under random mating and is almost certainly an artifact

## 7. Compute the inbreeding coefficient (f) for each SNP, and make a histogram of f. You can use function HWf for this purpose. Give descriptive statistics (mean, standard deviation, etc) of f calculated over the set of SNPs. What distribution do you expect f to follow theoretically? Use a probability plot to confirm your idea.
```{r}

#  F for all SNPs
f_values <- HWf(genotype_counts)

# Hist
hist(f_values, breaks = 50, col = "steelblue",
     main = "Distribution of F across SNPs",
     xlab = "F", ylab = "Number of SNPs")
abline(v = 0, col = "red", lty = 2)

# Descriptive statistics
cat("Mean:", round(mean(f_values, na.rm = TRUE), 4), "\n")
cat("SD:", round(sd(f_values, na.rm = TRUE), 4), "\n")
cat("Median:", round(median(f_values, na.rm = TRUE), 4), "\n")
cat("Min:", round(min(f_values, na.rm = TRUE), 4), "\n")
cat("Max:", round(max(f_values, na.rm = TRUE), 4), "\n")

# QQ plot
qqnorm(f_values, main = "QQ plot of F values", pch = 19, col = "steelblue")
qqline(f_values, col = "red", lwd = 2)
```

Some SNPs show extreme $F$ values, suggesting real deviations from HWE 

## 8. Apply the exact test for HWE to each SNP, using different significant levels. Report the number and percentage of significant variants using an exac test for HWE with $alpha$= 0.10, 0.05, 0.01 and 0.001. State your conclusions.
```{r}

# exact test to all SNPs
hwe_results <- HWExactStats(genotype_counts)

#  significance levels
alphas <- c(0.10, 0.05, 0.01, 0.001)

#  significant SNPs at each level
results <- data.frame(
  alpha = alphas,
  n_significant = sapply(alphas, function(a) sum(hwe_results < a, na.rm = TRUE)),
  pct_significant = sapply(alphas, function(a) mean(hwe_results < a, na.rm = TRUE) * 100)
)

#  expected by chance
results$expected_chance <- results$alpha * length(hwe_results)


print(results)
```

